---
title: "Human Activity Recognition Project"
author: "Massimo Morelli"
date: "Wednesday, June 18, 2014"
output: html_document
---

## Summary
In this report I try to train a machine learning model to predict the activity of the subject in a HAR (Human Activity Recognition ) excercise. HAR is a growing field of research, which tries to understand Human Behavior from observations regarding the subject and its environment. With a Random Forest predictor the Accuracy of TBD is reached, even with modeste computing resource. 

## Method

### Data collection
Data are made available from the group Groupware@LES ([1]), and comprise a a data set with TBD observation for training and cross validation, and a data set wih 20 data without the variable to predict (variable "classe")
```{r}
print("load data phase")
```

### Exploratory analysis
Data are loaded in a dataframe. There are 160 features, and many of them have a great proportion of NA. SThere appear to be a strong collinearity in many of the variables, so some kind of dimension reduction appear mandatory. Some column appear to be related to the experiment (windows variable, timestam variable, X) and even if in the preliminary analysis performed they seems to be very helpful in predict the variable of interest, I don't think that would be fair to introduce them in the analysis, so I excluded them.

### Statistical modeling
I divided the training data in three set: training, test, validation, to be used to training the model, to test various models, to predict out of sample accuracy. 

Due to the scarce computing resource at my disposal I kept the train set extremely small. Once the model is chosen I retrained it with more data (TBD). 

I started with reduction modeling. I removed all the variable with 
A Random Forest model has been trained with the help of the R package "Caret". Other models have much lowwer accuracy TBD

### Reproducibility
All analyses performed in this manuscript are reproduced in the R markdown file HAR_Finale.Rmd 


[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 

[5] R Markdown Page. URL: http://www.rstudio.com/ide/docs/authoring/using_markdown. Accessed 18.06.2014

===


```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
